---
title: "Raster resampling"
output: github_document
---

This is a deep dive on raster resampling - an important aspect of spatial analyses. There are a number of ways that rasters can be resampled and the method you use is dependent on the data.

Let's load up the data we need. For this example, we are going to load up 3 different rasters of data from Burkina Faso. Elevation (continuous outcome), land use (categorical) and population (count - from the [WorldPop project](https://www.worldpop.org/))

```{r}
library(raster)
BF_elev <- raster::getData("alt", country="BF")
BF_land_use <- raster("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week2/Lab_files/BF_land_use.tif")
pop <- raster("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week2/Assignment/BF_pop.tif")
```

In this case, `BF_land_use` isn't in the same projection as `BF_elev` and `pop`, so let's reproject. Remember that as land use is categorical, we have to specify `method = "ngb"` so that nearest neighbour interpolation is used during reprojection, otherwise it will try to use binlinear interpolation which will result in meaningless values.

```{r}
crs(BF_land_use)
crs(BF_elev)
BF_land_use <- projectRaster(BF_land_use, crs=crs(BF_elev), method="ngb") # reproject - can time a little time
```

For demo purposes, we are going to crop the rasters across an arbitray bounding boox. 
```{r}
BF_elev_crop <- crop(BF_elev, extent(BF_elev,499,507, 301,309))
pop_crop <- crop(pop, extent(BF_elev_crop))
BF_land_use_crop <- crop(BF_land_use, extent(BF_elev_crop))
```

Now let's create a dummy lower resolution raster we are going to resample to
```{r}
new_raster <- aggregate(BF_elev_crop, fact=3)
new_raster <- raster::shift(new_raster, dx=0.004, dy=0.002)
new_raster_outline <- rasterToPolygons(new_raster, dissolve=TRUE)
```

Let's plot the cropped elevation raster and overlay the grid of the new raster we are resample to
```{r}
plot(BF_elev_crop)

# plot cell outlines
lines(new_raster_outline)
```

Now let's resample elevation to the new raster. Here we can specify `method = binlinear` because this type of interpolation makes sense for a continuous outcome like elevation. In fact `method = binlinear` is the default so we don't really need to specify.
```{r}
BF_elev_crop_resampled_bilin <- resample(BF_elev_crop, new_raster, method="bilinear")
plot(BF_elev_crop_resampled_bilin)
lines(new_raster_outline)
```

Now let's repeat the resampling process with the land use raster. First let's plot and take a look at the frequency of different land use categories.
```{r}
plot(BF_land_use_crop)
BF_land_use_crop
table(BF_land_use_crop[])
```


As this is categorial, and given that new raster is coarser (lower resolution), first aggregate using the mode (i.e. most frequent) class. He we aggregate by a factor of 9 because the resolution of `the`new_raster` is 9 times coarser thanBF_land_use_crop
```{r}
BF_land_use_crop_aggregated <- aggregate(BF_land_use_crop, fun='modal', fact = 9)
plot(BF_land_use_crop_aggregated)
lines(new_raster_outline)
```

Now that the raster is at the right resolution, let's resample to the same grid
```{r}
BF_land_use_crop_aggregated_resamp <- resample(BF_land_use_crop_aggregated, new_raster, method = "ngb")
plot(BF_land_use_crop_aggregated_resamp)
lines(new_raster_outline)
```

Now we have resampled a continuous and categorical raster, let's repeat on population. The raw data in the `pop_crop` shows the number of individuals per cell. If we want to resample to a coarser resolution, we are probably most interested in maintaining these as counts and therefore we want to sum the numbers as opposed to use interpolation or nearest neighbour.

Let's have a look at the population raster
```{r}
plot(pop_crop, col=topo.colors(64))
cellStats(pop_crop, sum) 
```

As for the land use example, the new raster is at lower resolution. Therefore have to aggregate first. As we want total population to remain the same, aggregate by summing population
```{r}
pop_crop_aggregated <- aggregate(pop_crop, fact = 3.012048, fun = sum)

# Check the total population is the same as the original raster
cellStats(pop_crop_aggregated, sum)

# Plot with new raster cell outlines
plot(pop_crop_aggregated, col=topo.colors(64))
lines(new_raster_outline)
```


Now that the raster is at the right resolution, we can use the resample function to estimate the counts in the new grid
```{r}
pop_crop_aggregated_resamp <- resample(pop_crop_aggregated, new_raster)
plot(pop_crop_aggregated_resamp, col=topo.colors(64))
lines(new_raster_outline)
```

The total should more or less be equal to the original - its not exact as we have to use interpolation to predict populations in the new cell. 
```{r}
cellStats(pop_crop_aggregated_resamp, sum) # Not bad..
```

