---
title: "Untitled"
output: github_document
---

This week we are going to explore methods to understand and predict risk across space from point data. These may be point level data (i.e. measurements of something of interest at particular points) or point process data (i.e. occurences of events in a given area). When you load this week's libraries, it may prompt you to download [XQuartz](https://www.xquartz.org/)

```{r, echo=F, message=F, warning=F}
library(mapview)
webshot::install_phantomjs()
```

```{r, message=FALSE, warning=FALSE}
library(Metrics)
library(spatstat)
library(raster)
library(sp)
library(geoR)
library(gtools)
library(lme4)
library(leaflet)
library(oro.nifti)
```

First load up some obfuscated malaria case-control data from Namibia. This is comprised of latitudes and longitudes of cases and controls.
```{r}
CaseControl<-read.csv("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week3/Lab_files/CaseControl.csv")
head(CaseControl)
```

To set ourselves up for further analyses, let's create objects of just cases and just controls
```{r}
#Create a new object with just the cases, recoded as a number 1
Cases<-CaseControl[CaseControl$case==1,]

#Create a new object with just the controls, recoded as a number 0
Controls<-CaseControl[CaseControl$case==0,]
```

We are also going to create a `SpatialPointsDataFrame` of the case-control data
```{r}
CaseControl_SPDF <- SpatialPointsDataFrame(coords = CaseControl[,c("long", "lat")],
                                           data = CaseControl[,c("household_id", "case")])
```

And get hold of a boundary file for Namibia
```{r}
NAM_Adm0<-raster::getData('GADM',country='NAM',level=0)
```

Let's plot and see what we have. First, create a color scheme based on the case classification (0 or 1)
```{r}
case_color_scheme <- colorNumeric(c("blue", "red"), CaseControl_SPDF$case)
```

Then, plot
```{r}
leaflet() %>% addTiles() %>% addCircleMarkers(data=CaseControl_SPDF, 
                                              color = case_color_scheme(CaseControl_SPDF$case),
                                              radius = 2)
```

## Risk Mapping using Kernel Density
To generate a kernel density estimate, we first need to generate point pattern object of points (aka ppp). First, we need to define a window defining the population from which the cases arose
```{r}
Nam_Owin <- owin(xrange=range(CaseControl$long),yrange=range(CaseControl$lat))
```

Now we can define the ppp object of the cases
```{r}
Cases_ppp <- ppp(Cases$long, Cases$lat, window = Nam_Owin)
plot(Cases_ppp)
```

We can now generate and plot a kernel density estimate of cases
```{r}
par(mar=c(rep(1,4)))
case_density <- density(Cases_ppp)
plot(case_density) # Units are intensity of points per unit square
```

Its possible to use different bandwidths. The larger the bandwidth, the smoother the density estimate. 
```{r}
par(mfrow=c(3,1), mar=c(1,1,1,1))
plot(density(Cases_ppp,0.02), main = "Bandwidth 0.02")
plot(density(Cases_ppp,0.1), main = "Bandwidth 0.02")
plot(density(Cases_ppp,bw.ppl), main = "Automatically selected bandwidth") # automatic bandwidth selection based on cross-validation
```

If you want to map using leaflet, you have to convert the density object to a rasterLayer with a coordinate reference system
```{r, warning=F, message=F}
# Create raster
density_raster <- raster(density(Cases_ppp, bw.ppl), crs = crs(NAM_Adm0))

# Plot
leaflet() %>% addTiles() %>% addRasterImage(density_raster, opacity=0.6)
```

But this is just a density of cases, e.g. it doesn't account for the denominator - the controls. To do this, we can use the kelsall & diggle method, which calculates the ratio of the density estimate of cases:controls

First we have to add 'marks' to the points. Marks are just values associated with each point such as case or control (1/0) 
```{r}
CaseControl_ppp <- ppp(CaseControl$long, CaseControl$lat, 
                       window = Nam_Owin, 
                       marks=as.factor(CaseControl$case))
```

Now we can use the `relrisk` function from the spatstat pakage to look at the risk of being a case relative to the background population. In order to obtain an output of relative risk, we must specify relative = TRUE in the code line (the probability of being a case, relative to probability of being a control). If the 'relative' argument is not included in the code line the argument is technically specified as 'FALSE' since this is the default and the output is the probability of being a case. You can set sigma (bandwidth), but the default is to use cross-validation to find a common bandwidth to use for cases and controls. See `?bw.relrisk` for more details.

```{r}
par(mar=rep(1,4))
risk_est <-  relrisk(CaseControl_ppp) 
plot(risk_est)
```

## Obtaining a relative risk of being a case

```{r}
par(mar=rep(1,4))
rel_risk_est <-  relrisk(CaseControl_ppp, relative = TRUE)
plot(rel_risk_est)
```

To plot on a web map, first specify the projection
```{r}
risk_raster <- raster(risk_est, crs = crs(NAM_Adm0))
```

Then define a color palette
```{r}
pal <- colorNumeric(palette=tim.colors(64), domain=values(risk_raster), na.color = NA)
```

Then plot with leaflet
```{r, warning=F, message=F}
leaflet() %>% addTiles("http://{s}.basemaps.cartocdn.com/light_all/{z}/{x}/{y}.png") %>% 
  addRasterImage(risk_raster, opacity=0.6, col = pal) 
```

## Interpolation of point (prevalence etc.) data
First load Ethiopia malaria prevalence data
```{r}
ETH_malaria_data <- read.csv("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week1/Lab_files/Data/mal_data_eth_2009_no_dups.csv",header=T)
```

Get the Ethiopia Adm 1 level boundary file using the raster package which provides access to GADM data
```{r}
ETH_Adm_1 <- raster::getData("GADM", country="ETH", level=1)
```

## Inverse distance weighting (IDW)
Inverse distance weighting is one method of interpolation. To perform IDW using the spatstat package, as per kernel density estimates, we have to create a ppp object with the outcome we wish to interpolate as marks. We have to start by setting the observation window. In this case, we are going to use the bounding box around Oromia State from which these data were collected. To set the window for the `ppp` function, we need to use the `owin` function. 

```{r}
oromia <- ETH_Adm_1[ETH_Adm_1$NAME_1=="Oromia",]
oromia_window <- owin(oromia@bbox[1,], oromia@bbox[2,])
```

Then define a ppp of the prevalence data
```{r}
ETH_malaria_data_ppp<-ppp(ETH_malaria_data$longitude,ETH_malaria_data$latitude,
                         marks=ETH_malaria_data$pf_pr,window=oromia_window)
```

Set the parameters for displaying multiple plots in one screen and plot different IDW results
NB: 1) power represents the power function we want to use
    2) 'at' can be 'pixels' where it generates estimates across a grid of pixels or 'points' where it interpolates values at every point using leave-one-out-cross validation
```{r}
par(mfrow=c(2,2))
plot(idw(ETH_malaria_data_ppp, power=0.2, at="pixels"),col=heat.colors(20), main="power = 0.2") 
plot(idw(ETH_malaria_data_ppp, power=0.5, at="pixels"),col=heat.colors(20), main="power = 0.5")
plot(idw(ETH_malaria_data_ppp, power=1, at="pixels"),col=heat.colors(20), main="power = 0.1")
plot(idw(ETH_malaria_data_ppp, power=2, at="pixels"),col=heat.colors(20), main="power = 2") 
```

To calculate the 'best' power to use, you can use cross-validation. This is possible using the argument at=points when running the idw function. There is no off the shelf function (that I know of) to do this, so you have to loop through different power values and find the one that produces the lowest error using cross-validation.
```{r}
powers <- seq(0.05, 2, 0.05) 
mse_result <- NULL 
for(power in powers){CV_idw <- idw(ETH_malaria_data_ppp, power=power, at="points") 
  mse_result <- c(mse_result, mse(ETH_malaria_data_ppp$marks,CV_idw)) 
  }
```
See which produced the lowest error
```{r}
optimal_power <- powers[which.min(mse_result)] 
plot(powers, mse_result) 
```
Plot observed versus expected with optimal power
```{r}
CV_idw_opt <- idw(ETH_malaria_data_ppp, power=optimal_power, at="points") 
plot(ETH_malaria_data_ppp$marks, CV_idw_opt, xlab="Observed prevalence",
     ylab="Predicted prevalence")
```

Plot using leaflet.

1. Convert to a raster
```{r}
ETH_malaria_data_idw_raster <- raster(idw(ETH_malaria_data_ppp, power=0.2, at="pixels"),
                                     crs= crs(ETH_Adm_1))
```
2. Define a color palette
```{r}
colPal <- colorNumeric(tim.colors(), ETH_malaria_data_idw_raster[], na.color = NA)
```
3. Plot
```{r, warning=F, message=F}
leaflet() %>% addTiles() %>% addRasterImage(ETH_malaria_data_idw_raster, col = colPal, opacity=0.7) %>%
  addLegend(pal = colPal, values = ETH_malaria_data_idw_raster[])
```

## Kriging
We are going to use the GeoR package to perform kriging. First, we have to create a geodata object with the package GeoR. This wants dataframe of x,y and data 
```{r}
ETH_malaria_data_geo <- as.geodata(ETH_malaria_data[,c("longitude","latitude","pf_pr")])
```

We can plot a summary plot using ther Lowes parameter. The Lowes option gives us lowes curves for the relationship between x and y
```{r}
plot(ETH_malaria_data_geo, lowes=T)
```

It's important to assess whether there is a first order trend in the data before kriging. We can see from the plots of the prevalence against the x and y coordinates that there isn't really any evidence of such a trend. Were there to be evidence, you can add `trend = 1st` or `trend = 2nd` to the plot command to see the result after havin regressed prevalence against x and y using a linear and polynomial effect respectively. 

Now generate and plot a variogram. As a rule of thumb, its a good idea to limit variogram estimation to half the maximum interpoint distance
```{r}
MaxDist <- max(dist(ETH_malaria_data[,c("longitude","latitude")]))  /2 
VarioCloud<-variog(ETH_malaria_data_geo, option="cloud", max.dist=MaxDist)
plot(VarioCloud) # all pairwise comparisons
```
To make it easier to interpret, we can bin points by distance
```{r}
Vario <- variog(ETH_malaria_data_geo, max.dist = MaxDist, trend = "2nd")
plot(Vario)
```

Its possible to change the way the variogram bins are constructed. Just be careful not to have too few pairs of points in any distance class.
NB: `uvec` agrument provides values to define the variogram binning (ie let's try bins of 0.2 decimal degrees, about 22 km)
```{r}
Vario <- variog(ETH_malaria_data_geo,max.dist=MaxDist,uvec=seq(0.01,MaxDist,0.2)) 
```

Let's look at the number in each bin
```{r}
Vario$n
```
What is the minimum? A rule of thumb is 30 in each bin
```{r}
min(Vario$n)
```
Plot
```{r}
plot(Vario,pch=16)
```

We can now fit variogram model by minimized least sqaures using different covariance models. 
In this case we are just going to use a 'spherical' and 'exponential' model. 

```{r}
VarioMod_sph<-variofit(Vario, cov.model = "sph")
VarioMod_exp<-variofit(Vario, cov.model = "exp")
```
Plot results
```{r}
plot(Vario,pch=16)
lines(VarioMod_sph,col="blue",lwd=2)
lines(VarioMod_exp,col="red",lwd=2) 
```

Get summaries of the fits
```{r}
summary(VarioMod_sph)
summary(VarioMod_exp)
```
In this case, the spherical model has a slightly lower sum of squares, suggesting it provides a better fit to the data.

Now we have a variogram model depicting the covariance between pairs of points as a function of distance between points, we can use it to Krig values at prediction locations. To allow us to compare with IDW, first get grid of points from the IDW example for comparison

1. Create prediction grid
```{r}
IDW <- idw(ETH_malaria_data_ppp, power=0.2, at="pixels")
pred_grid_x <- rep(IDW$xcol,length(IDW$yrow))
pred_grid_y <- sort(rep(IDW$yrow,length(IDW$xcol)))
pred_grid <- cbind(pred_grid_x,pred_grid_y)
```
2. Now krig to those points
```{r}
KrigPred <- krige.conv(ETH_malaria_data_geo, loc=pred_grid,
                       krige=krige.control(obj.model=VarioMod_sph))
```
Visualize predictions
```{r}
image(KrigPred,col=heat.colors(50))
```

If you want to create a raster of your predictions, you can use the `rasterFromXYZ` function
```{r}
KrigPred_raster <- rasterFromXYZ(data.frame(x=pred_grid_x,
                                 y=pred_grid_y,
                                 z=KrigPred$predict))

plot(KrigPred_raster)
points(ETH_malaria_data[,c("longitude","latitude")],
       cex = ETH_malaria_data$pf_pr * 10)
```

Generating cross-validated predictions in straightforward in geoR using the [`xvlalid`](https://www.rdocumentation.org/packages/geoR/versions/1.8-1/topics/xvalid) function. Two types of validation are possible:
1. leaving-on-out cross validation where each data location (all or a subset) is removed in turn and predicted using the remaining locations, for a given model.
2. External validation which can predict to locations outside of the dataset
Here we will use the default leave-one-out xvalidation for all points
```{r}
xvalid_result <- xvalid(ETH_malaria_data_geo, model = VarioMod_sph)

# Plot on log odds scale
plot(xvalid_result$data,xvalid_result$predicted, asp=1,
     xlab = "Observed prevalence", ylab="Cross-validated predicted prevalence")
abline(0,1)
```

You might notice that some of the kriged values are <0. As we are modeling probabilities this can't be true. In these situations, it is possible to apply a transformation to your data before kriging and then back-transform results. One transformation useful for probabilities is the logit transform (used in logistic regression). The `logit` and `inv.logit` functions from the package `gtools` can be used for this. Note that it doesn't work if you have 0 values as you can't log(0). You can add a small amount to avoid this situation. The process would look like this

```{r}
# Add small amount to avoid zeros
ETH_malaria_data$pf_pr_adj <- ETH_malaria_data$pf_pr + 0.001

# Apply logit transform and convert to geodata
ETH_malaria_data$pf_pr_logit <- logit(ETH_malaria_data$pf_pr_adj)
ETH_malaria_data_geo_logit <- as.geodata(ETH_malaria_data[,c("longitude","latitude","pf_pr_logit")])

# Fit (spherical) variogram
Vario_logit <- variog(ETH_malaria_data_geo_logit, max.dist = MaxDist)
VarioMod_sph_logit <- variofit(Vario_logit, cov.model = "sph")

# Get CV kriged predictions
xvalid_result_logit <- xvalid(ETH_malaria_data_geo_logit, model = VarioMod_sph_logit)
xvalid_result_inv_logit <- inv.logit(xvalid_result_logit$predicted)
```


## Pop quiz
* How could you compare how well the best fitting IDW performs versus kriging? 
* Which appears to be more accurate?
* Can you visualize where predictions from IDW differ to kriging?
* Does inclusion of a trend surface improve kriging estimates?

Answers [here](https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/_posts/week_3_cheat_sheet.R)

## Key readings

#### Good overview
Pfeiffer, D., T. P. Robinson, M. Stevenson, K. B. Stevens, D. J. Rogers and A. C. Clements (2008). Spatial analysis in epidemiology, Oxford University Press Oxford. Chapter 6. 

#### Technical paper covering kernel estimation of relative risk. Key reference but not necessary to understand in detail.
Kelsall, Julia E., and Peter J. Diggle. "Kernel estimation of relative risk." Bernoulli 1.1-2 (1995): 3-16.

#### Illustration of the Kelsall Diggle approach used to map sleeping sickness risk
[Simarro, Pere P., et al. "Estimating and mapping the population at risk of sleeping sickness." PLoS neglected tropical diseases 6.10 (2012): e1859.](https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0001859)


## Additional readings
#### Nice example of kriging applied across space and time
[Gething, Peter W., et al. "A local space–time kriging approach applied to a national outpatient malaria data set." Computers & geosciences 33.10 (2007): 1337-1350.](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.0030271)

#### Additional example of Kelsall Diggle in action
Di Salvo, Francesca, et al. "Spatial variation in mortality risk for hematological malignancies near a petrochemical refinery: A population-based case-control study." Environmental research 140 (2015): 641-648.