---
layout: post
title: Week 1 - Visualizing spatial data
featured-img: unemployment_USA_2016
output: github_document
always_allow_html: yes
---

Week 1 - visualizing spatial data
---------------------------------

Welcome to week 1 of spatial analysis for public health. This week we will be learning about the process of moving from visualizing spatial data through to exploration and analysis. We will get our hands dirty with some R code and learn how to make beautiful maps. This week's [lecture]() will focus on some of the concepts behind spatial epidemiology. The code below covers loading and visualizing spatial data in R. A more complete walkthrough of this code can be seen in [this video](). You will then have a chance to apply that code to new data and questions in this week's [assignment]().


Lab 1: Working with Spatial Data in R
-------------------------------------

The simplest data is a table with coordinates (i.e. point data). For this assignment, we'll work with malaria prevalence point data from Ethiopia. These data were downloaded from the Malaria Atlas Project data repository and were originally collected as part of [a study](https://link.springer.com/article/10.1186/1475-2875-10-25) conducted in 2009. 

First get the necessary libraries for this week

```{r}
library(sp)
library(raster)
library(rgdal)
library(leaflet)
```

Import the data

```{r}
ETH_malaria_data <- read.csv("https://raw.githubusercontent.com/HughSt/HughSt.github.io/master/course_materials/week1/Lab_files/Data/mal_data_eth_2009_no_dups.csv",
                            header=T)
```

The columns should be self-explanatory, but briefly: 
* examined = numbers tested
* pf_pos = of those tested, how many were positive for Plasmodium falciparum malaria 
* pf_pr = Plasmodium falciparum parasite rate which is the same as infection prevalence or proportion infected (i.e. pf_pos / examined)
* longitude = longitude of school in decimal degrees 
* latitude = latitude of schoolin decimal degrees

``` {r}
# Look at the first few rows
head(ETH_malaria_data)

# Create a histogram of the prevalence
hist(ETH_malaria_data$pf_pr, breaks=20)
```

# Plotting and mapping spatial data
It is possible to use R's base graphics to plot points, treating them like any other data with x and y coordinates. For example, to get a plot of the points alone
```{r}
plot(ETH_malaria_data$longitude, ETH_malaria_data$latitude,
     ylab = "Latitude", xlab="Longitude") #boring!
```


You might want to vary the size of the circle as a function of a variable. For example, if we wanted to plot points with size relative to prevalence we can use the expansion argument `cex`
```{r}
# Use the cex function to plot circle size as a function of a variable. In this case prevalence. As the values are very small we can multiply by an arbitrary amount for visualization purposes
plot(ETH_malaria_data$longitude, ETH_malaria_data$latitude,
     cex = ETH_malaria_data$pf_pr * 10,
     ylab = "Latitude", xlab="Longitude")
```


## Working with 'Spatial' objects
In R, it is sometimes useful to package spatial data up into a 'Spatial' class of object using the `sp` package. This often makes it easier to work with and is often a requirement for other functions. The [sp](https://cran.r-project.org/web/packages/sp/index.html) package allows you to put your data into specific spatial objects, such as `SpatialPoints` or `SpatialPolygons`. In addition, if your data are more than just the geometry, i.e. if you have data associated with each spatial feature, you can create spatial DataFrames, i.e. `SpatialPointsDataFrames` and `SpatialPolygonsDataFrames`. For example, if we wanted to create a SpatalPointsDataFrame using the Ethiopia data:

```{r}
ETH_malaria_data_SPDF <- SpatialPointsDataFrame(coords = ETH_malaria_data[,c("longitude", "latitude")],
                                      data = ETH_malaria_data[,c("examined", "pf_pos", "pf_pr")],
                                      proj4string = CRS("+init=epsg:4326")) # sets the projection to WGS 1984 using lat/long. Optional but good to specify

# Summary of object
ETH_malaria_data_SPDF

# SPDFs partition data elements, e.g. the coordinates are stored separately from the data
head(ETH_malaria_data_SPDF@coords)
head(ETH_malaria_data_SPDF@data)

# You can quickly access the data frame as per a standard data frame, e.g.
head(ETH_malaria_data_SPDF$pf_pr)

# You can use the plot or spplot function to get quick plots
plot(ETH_malaria_data_SPDF)
spplot(ETH_malaria_data_SPDF, zcol = "pf_pr")
```

Let's have a look at SpatialPolygonsDataFrames. To load a polygon shapefile (or other file types), you can use the `readOGR` function from the `rgdal` package. For example, if you wanted to load in the province boundaries for Ethiopia shapefile `ETH_Adm_1` from the `ETH_Adm_1_shapefile` folder on [GitHub](https://github.com/HughSt/HughSt.github.io/tree/master/course_materials/week1/Lab_files/Data/ETH_Adm_1_shapefile), assuming you have downloaded the folder of files you would use the following command

```
ETH_Adm_1 <- readOGR("ETH_Adm_1_shapefile", "ETH_Adm_1")
```

As it happens, admin boundary data is accessible using the `getData` function from the `raster` package. Be careful as some other packages also have a `getData` function, so to specify that you want to use the `getData` function from the `raster` package you can use the following code

```{r}
# You first need the ISO3 codes for the country of interest. You can access these using `ccodes()`. For Ethiopia, the ISO3 is ETH

# The getData function then allows you to retrieve the relevant admin level boundaries from GADM.
ETH_Adm_1 <- raster::getData("GADM", country="ETH", level=1) 
```

Now we can plot the point data in context
```{r}
# Plot both country and data points
plot(ETH_Adm_1)
points(ETH_malaria_data$longitude, ETH_malaria_data$latitude,
     cex = ETH_malaria_data$pf_pr * 10,
     ylab = "Latitude", xlab="Longitude",
     col="red")
```


# Plotting data using web maps
Rather than just relying on R base graphics, we can easily create webmaps using the `leaflet` package. There are many basemaps available. See [here](https://leaflet-extras.github.io/leaflet-providers/preview/). For any map, identify the Provider name, e.g. "OpenStreetMap.Mapnik", by clicking on the map.

```
# Define your basemap
basemap <- leaflet() %>% addTiles()
basemap

# Or choose another basemap
basemap <- leaflet() %>% addProviderTiles("Esri.WorldImagery")
basemap

#Let's choose a simple one
basemap <- leaflet() %>% addProviderTiles("CartoDB.Positron")
```

You can use the 'piping' command %>% to add layers. As our point and polygon data are already 'Spatial' object this is easy

```
basemap %>% addPolygons(data=ETH_Adm_1)


# to change the colors/line weight
basemap %>% addPolygons(data=ETH_Adm_1, color = "red", 
                        weight = 1, fillOpacity = 0.2)

#You can also add popups
basemap %>% addPolygons(data=ETH_Adm_1, 
                        popup = ETH_Adm_1$NAME_1)

# If you want to add points as well
basemap %>% addPolygons(data=ETH_Adm_1, weight = 2,
                        popup = ETH_Adm_1$NAME_1) %>%
  
            addCircleMarkers(data=ETH_malaria_data_SPDF,
                             color="red", radius = 2)
```

The leaflet package also has some nice functions for generate color palettes that map to a variable you want to display. For example, if we wanted to create a color ramp relative to prevalence we could use the `colorQuantile` function. See ?colorQuantile for other ways to build color palettes such as `colorNumeric`
```
library(wesanderson) # for a nice color palette
colorPal <- colorNumeric(wes_palette("Zissou1")[1:5], ETH_malaria_data_SPDF$pf_pr, n = 5)

# colorPal is now a function you can apply to get the corresponding
# color for a value
colorPal(0.1)

basemap %>% addPolygons(data=ETH_Adm_1, weight = 2, fillOpacity=0,
                        popup = ETH_Adm_1$NAME_1,
                        color = "gray") %>%
  
  addCircleMarkers(data=ETH_malaria_data_SPDF,
                   color = colorPal(ETH_malaria_data_SPDF$pf_pr), 
                   radius = 2,
                   popup = as.character(ETH_malaria_data_SPDF$pf_pr))
```

You might want to add a legend. This just goes on as another layer on the map. First define the labels. In this case, we are using quintiles.
```
basemap %>% addPolygons(data=ETH_Adm_1, weight = 2, fillOpacity=0,
                        popup = ETH_Adm_1$NAME_1,
                        color = "gray") %>%
  
  addCircleMarkers(data=ETH_malaria_data_SPDF,
                   color = colorPal(ETH_malaria_data_SPDF$pf_pr), 
                   radius = 2,
                   popup = as.character(ETH_malaria_data_SPDF$pf_pr)) %>%
  
  addLegend(pal = colorPal, 
            title = "Prevalence",
            values = ETH_malaria_data_SPDF$pf_pr )
```


For more complex popups, you can define the HTML
```
basemap %>% addPolygons(data=ETH_Adm_1, weight = 2, fillOpacity=0,
                        popup = ETH_Adm_1$NAME_1,
                        color = "gray") %>%
  
  addCircleMarkers(data=ETH_malaria_data_SPDF,
                   color = colorPal(ETH_malaria_data_SPDF$pf_pr), 
                   radius = 2,
                   popup = paste("<p>","Prevalence:",
                                 round(ETH_malaria_data_SPDF$pf_pr,2),
                                 "<p>")) %>%
                                 
 addLegend(pal = colorPal, 
            title = "Prevalence",
            values = ETH_malaria_data_SPDF$pf_pr )
```


# Plotting raster data
If you have a local raster file (e.g. a .tif file), you can use the `raster` command to load the file into R. For example, if you download the ["elev_ETH.tif"](https://github.com/HughSt/HughSt.github.io/blob/master/course_materials/week1/Lab_files/Data/elev_ETH.tif?raw=true) file from this week's data, you would use the following to load from a local source:

```
elev <- raster("elev_ETH.tif")
```

You can also load using a URL, e.g.

```
elev <- raster("https://github.com/HughSt/HughSt.github.io/blob/master/course_materials/week1/Lab_files/Data/elev_ETH.tif?raw=true")
```

The `getData` functon from the `raster` package allows you to get hold of some select raster data, such as elevation and bioclimatic layers. To get hold of elevation for Ethiopia, use the following

```{r}
elev <- raster::getData("alt", country="ETH")
elev
```

You can plot using the plot function
```{r}
plot(elev)
```

Alternatively, you can use leaflet
```
basemap %>% addRasterImage(elev)
```

If you want to add a legend, you have to define the color palette first
```
# Define palette
raster_colorPal <- colorNumeric(topo.colors(64), values(elev), na.color = NA)

# Plot
basemap %>% addRasterImage(elev, color = raster_colorPal) %>%
addLegend(values = values(elev), pal = raster_colorPal)
```

If you want to export the data, there are several options.

1. Export button in the Viewer pane. Using 'Save as webpage' creates an html file which you can open using a browser.

2. Save as kml for someone to open in Google Earth
```
library(plotKML)
plotKML(ETH_malaria_data_SPDF) # see ?plotKML for more options
```

# Pop quiz

When using the base graphics to plot point data, how might you make the points larger, but still relative to prevalence?


Resources
---------

The R packages [sp](https://cran.r-project.org/web/packages/sp/index.html) and [raster](https://cran.r-project.org/web/packages/raster/index.html) are both important packages for spatial analysis.

R studio also have a fantastic [site](https://rstudio.github.io/leaflet/) outlining the use of leaflet


